Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-09-11 jss/EpiCompare-paper/EpiCompare-paper.tex

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process, as has been seen in the current COVID-19 pandemic
\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.

With many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through the
MIDAS (Models of Infectious Disease Agent Study) network
\citep{midasnetwork2021} available, this field has a number of devices
to \textcolor{orange}{\sout{aid}}
\textcolor{orange}{help}\footnote{\textcolor{violet}{[Ben says: I'm not sure this an improvement]}}
an individual practitioner decide the correct approach. For example,
\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
\pkg{pomp} have all made significant steps in standardizing the flow of
the data analysis pipeline for epidemic modeling through digitizing data
sets, making accessible statistical models, and providing a plethora of
educational material for both coding novices and experts alike
\citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific
portion of the analysis pipeline. These modeling tools usually require
learning package-specific syntax and often don't provide easy ways to
compare and assess their models on new data. Moreover, exploring
modeling and comparing epidemics require transforming and
\textit{tidying} data in different ways. To fill these gaps, we present
our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
is to aid and advance research in the area of comparison and assessment
of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we
illustrate the data analysis pipeline of infectious diseases as 1) data
pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
simulating, 4) post-processing, and 5) comparison and assessment; where
each previous part of the pipeline influences the next. \pkg{EpiCompare}
provides tools to aids practitioners in all areas of this pipeline.

@@

One of \pkg{EpiCompare}'s main contribution to comparison and assessment
of epidemics is through tools that provide \textit{time-invariant}
assessments. Epidemics, despite being defined as a process that evolves
over time, often need to be compared in a way not constrained to initial
times or time scales in order to understand the processes at play. With
time-invariant analysis, comparing decades-long outbreaks of HIV in the
US to a 10 day outbreak of norovirus on a cruise ship is possible.
Compared to time-dependent comparison tools for state-space modeling,
time-invariant analysis can make it easier to compare state-space
epidemic representations in a more global, holistic fashion. Many
time-dependent comparison tools only examine the proportion of
individuals in each state (at a given time) in a piece-wise / marginal
fashion. These time-dependent approaches can reduce the amount of
connections that can be seen and insights that can be drawn\}, similar
to examining projections of a multidimensional distribution onto a
single axis, one at a time. Tools in \pkg{EpiCompare} extend the user
toolkit to evaluate epidemics within a time-invariant lens. At the same
time, the goal of \pkg{EpiCompare} is not to supplant existing
infectious disease modeling tools and software but, rather, is a
concerted effort to create standard and fair comparisons among models
developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a demonstration of the tools through a
detailed
example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer? up to you} \textcolor{violet}{The current suggestion isn't clearer - feel free to try again.}}

\hypertarget{sec:time-invariant}{%
\section{Time-invariant analysis}\label{sec:time-invariant}}

\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we highlight some weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we first demonstrate the inability of time-dependent tools to
adequately quantify a classic epidemic parameter, the reproduction
number \(R_0\). Then we motivate new time-invariant approaches for more
complex situations where \(R_0\) does not capture the complexities of
outbreaks. This leads to the final subsection that defines ways to view
epidemics in a time-invariant lens and discusses natural properties of
these representations that allow for clearer ways to compare models and
epidemics.

\subsection[Motivation through the reproduction number $R_0$]{Motivation
of time-invariant analysis through the reproduction number
\(R_0\)}\label{sec:r0_subsection}

Epidemiological research is often interested in understanding intrinsic
properties of the epidemic, independent from when the epidemic occurred
or how frequently data was collected. One of the most famous numerical
summaries of an epidemic is the reproduction number - \(R_0\). This
numerical summary is a time-invariant value that is defined as the
expected number of infections caused by a single infect or who is added
to a completely susceptible population \citep{anderson1992}.
\citet{Gallagher2020} showed that the estimation of \(R_0\) can be
sensitive to time-dependent parameters like estimation of the beginning
and end of an epidemic. This time-dependent sensitivity in estimating
\(R_0\) reflects a general problem in epidemiology surrounding how to
transform time-dependent information into insights into desirable,
intrinsic properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an
epidemic than just the reproduction number. One common time-dependent
tool used to better understand epidemics is a series of time series line
plots that track the proportion of the population in a given state
(e.g.~infected) at a given time. Figure
\ref{fig:different-scales-standard} visualizes two different simulated
epidemics with population states (S)usceptible, (I)nfectious and
(R)ecovered. These simulations have been generated under a discrete
approximation of \citet{Kermack1927}'s SIR model. The Kermack and
McKendrick model captures the transitions from one state to the next as
a system of ordinary differential equations

\begin{align}\label{eq:sir-ode}
    S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
    I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
    R^\prime(t) &= \gamma I(t) \nonumber,
\end{align}

where \(N\) is the total number of individuals, \(\beta\) is the rate of
infection, and \(\gamma\) is the rate of recovery.

From this model, the reproduction number is the ratio of the infection
rate to the recovery rate, \(R_0 = \beta/\gamma\).

@@

\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ parameters but the same initial reproduction number $R_0$ = 2.  Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
\end{figure}
\end{CodeChunk}

Even though Figure \ref{fig:different-scales-standard}'s visual analysis
might be able to present more complex properties of the epidemic, one
might wish to understand how these two simulated epidemics' \(R_0\)
values compare. Even though these two simulated epidemics appear
different in these figures (including having different infection peaks),
there is no real intuitive way to compare the epidemics' \(R_0\) values.
In fact, both of these simulated epidemics started with the same
population (1000 total people with 10 infected) and have the same
\(R_0\) value, just with slightly different infection and recovery rates
(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
this time-dependent visualization cannot be used to compare a very
simple but powerful numerical summary, \(R_0\).

\hypertarget{ternary-plots-a-time-invariant-visualization-tool}{%
\subsection{Ternary plots: a time-invariant visualization
tool}\label{ternary-plots-a-time-invariant-visualization-tool}}

The faceted time series plot, like that seen in Figure
\ref{fig:different-scales-standard}, not only fails to allow the
practitioner to compare simulated epidemics' \(R_0\) values but also
presented epidemics' trajectory data so that only the marginal
information for each state can be examined at a given time. A more
holistic approach to visualizing the overall path of our simulated
epidemics is to examine how each epidemic traverses the
three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
just that, presenting the trajectories of the simulated epidemics in
this three-dimensional space. For state space models like in our
example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
to \(N\) (the total population size) we can visual these point in a a
two-dimensional \textit{ternary} plot, as seen in Figure
\ref{fig:different-scales-tern}'s right plot. In both of Figure
\ref{fig:different-scales-tern}'s plots we observe that the two
epidemics are on the same trajectory. In this simple generative example
setting, being on the same trajectory indicates that the two epidemics
have the same \(R_0\) value. This can be proven mathematically.

@@

\caption{\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.  Right: the gray-shaded region and epidemic trajectories shown from (left) now shown in two-dimensional space.  This is more commonly known as a ternary plot.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\begin{theorem}\label{thm:sir-scale}
 Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e. $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$. Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\), then every point on the epidemic path of the
first SIR model can be mapped to a point on the epidemic path of the
second SIR model. In other words, the two epidemics form the same
filamental trajectory.

\hypertarget{time-invariance-beyond-sir-models-trajectories-and-filaments}{%
\subsection{Time-invariance beyond SIR models: Trajectories and
Filaments}\label{time-invariance-beyond-sir-models-trajectories-and-filaments}}

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space allows us
to better compare the overall structure of the epidemic and see how the
population is directly impacted. In this section we present
time-invariant tools that can be applied to complex epidemics where the
epidemic's generative process is unknown and can have more than three
states. These tools leverage the idea that an epidemic can be viewed as
a trajectory and that many properties of the epidemic are well captured
when we do so. This approach is useful when the epidemic of interest has
only gone through a single realization of its outbreak (before the
population of individuals become susceptible again).

The first set of tools allows a practitioner to define distances between
epidemics whose time features do not align. For completed epidemics, one
way to better examine their properties is to represent their filamental
trajectories as a finite sequence of equally spaced points. This
representation induces a natural distance between epidemics,
specifically:

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (with
\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
time-invariant, and can be plugged into multiple distance-based
assessment tools to examine the overall ``extremeness'\,' of points,
including pseudo-density estimators and depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be useful when comparing the true epidemic to a set of
simulated epidemics, and practitioners can interpret the epidemic's
extremeness score relative to the extremeness scores of the simulations
very easily.

In addition to being used to better understand completed epidemics,
time-invariant tools can be used to aid prediction of future epidemics
by providing a state-space regions in which we expect the true epidemic
to traverse. In settings where the epidemic only generates a single
outbreak, these regions can be very telling if simulation models capture
the epidemic's structure. In \pkg{EpiCompare} we create geometric
prediction regions around all but the \(\alpha\) proportion of most
extreme simulated trajectories. These geometric regions can also be used
to compared simulation models that have different time scales,
parameters and even different statistical philosophies, through set
different distances. Although visualization is the easiest when the
epidemic has three states, prediction regions can be useful to assess
and compare simulation models where the epidemics have multiple states.

Overall, there are many tools to aid in the assessment and comparison of
epidemics and models that avoid being affected by time-based parameters.
We believe the time-invariant analysis provides many insights and should
be in the toolkit of many epidemiologists. \pkg{EpiCompare} provides a
strong starting point to do just that.

>>>>

The recent (and on-going) COVID-19 global pandemic has galvanized public
interest in understanding more about infectious disease modeling and has
highlighted the usefulness of research in the area of infectious disease
epidemiology. Infectious diseases inflict enormous burdens on the world:
millions of lives lost and trillions of dollars spent yearly. Infectious
disease models typically attempt to do one or more of the following: 1)
predict the spread of current and future epidemics
\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
and visualizations from universities, many branches and levels of
government, and news organizations are an important first step of the
process
\textcolor{violet}{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}

With the many visualization and exploratory tools, models and modeling
paradigms, and reviews and comparisons in the literature and through the
MIDAS (Models of Infectious Disease Agent Study) network
\citep{midasnetwork2021}, this field has a lot of devices to aid an
individual practitioner decide the correct approach. For example,
\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
\pkg{pomp} have all made significant steps in standardizing the flow of
the data analysis pipeline for epidemic modeling through digitizing data
sets, making accessible statistical models, and providing a plethora of
educational material for both coding novices and experts alike
\citep{surveillance2017,Jenness2018,King2016}.

At the same time, analysis packages often only address a specific
portion of the analysis
pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
\textcolor{violet}{These m}odeling
tools\textcolor{violet}{\sout{, which}} usually require learning
package-specific syntax\textcolor{violet}{\sout{,}} and often don't
provide easy ways to compare and assess their models on new data.
Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
\textcolor{violet}{and comparing} epidemics require transforming and
\textit{tidying} data in different ways. To fill these gaps, we present
our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
is to aid and advance research in the area of comparison and assessment
of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we
illustrate the data analysis pipeline of infectious diseases as 1) data
pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
simulating, 4) post-processing, and 5) comparison and assessment; where
each previous part of the pipeline influences the next. \pkg{EpiCompare}
provides tools to aids practitioners in all areas of this pipeline.

@@

\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way. Epidemics, despite by definition being a
process that evolves over time, often need to be compared in a way not
constrained to initial times or time scales to understand the processes
at play.
\textcolor{violet}{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion. \sout{Moreover, m} M}any
current \textcolor{violet}{time-dependent} comparison tools for
state-space models (e.g.~SIR models)
\textcolor{violet}{\sout{highlight} examine} the proportion of
individuals in each state (at a given time) in a piece-wise / marginal
fashion. \textcolor{violet}{These \sout{This}}
approach\textcolor{violet}{es} may reduce the amount of connections that
can be seen, similar to projections of a multidimensional distribution
onto a single axis at a time. Tools in \pkg{EpiCompare} give the user
the ability to extend their toolkit to evaluate epidemics within a
time-invariant lens. The goal of \pkg{EpiCompare} is not to supplant
existing infectious disease modeling tools and software but, rather, is
a concerted effort to create standard and fair comparisons among models
developed for disease outbreaks and outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
demonstrating of the tools through a detailed example of a full data
analysis pipeline.

\section[Time-invariant]{Motivation and tools for time-invariant
analysis}\label{sec:time-invariant}

\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
a global, not marginal view of how epidemics move through populations
and (2) by treating full epidemics as filamental trajectories. The
following section aims to highlight the strengths of
\textit{time-invariant} \textcolor{orange}{analysis} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

Mathematically, epidemics are complex objects. They can be hard to
assess and compare to one another due to the differences in the
diseases, the location where the outbreak occurs, how the affected
population reacts, and the time
\sout{related}\textcolor{orange}{related} features (including start of
the epidemic, speed of infection and more). Time-invariant analysis
makes different epidemics easier to compare by removing many time
dependent aspects of an epidemic. \sout{Instead,}
\textcolor{orange}{Time-invariant analysis} focuses
\textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
\textcolor{violet}{[Ben wants to try this sentence again.]}

\subsection[motivating through R0]{Motivating time-invariant analysis
through the reproduction number \(R_0\)}\label{sec:r0}

Time-invariant analysis, as it appears in \pkg{EpiCompare},
\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
comparing different epidemics. With time-invariant analysis, comparing
the decades-long outbreak of HIV in the US to a 10 day outbreak of
norovirus on a cruise ship is \sout{still} possible. Time-dependent
problems can arise when estimating epidemiological parameters, including
the reproduction number \(R_0\).
\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence.}
  \textcolor{violet}{I still think it adds value to the story and I'm not sure people really read section titles that are long.}}

\(R_0\) is probably the most famous \sout{time-invariant} numerical
summary of an epidemic and is often associated with the
Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
\(R_0\) is \sout{a one-number summary of a disease and is }defined as
the expected number of infections caused by a single infector who is
added to a completely susceptible population \citep{anderson1992}.
\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{. Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a difficult quantity to estimate.}
\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
To demonstrate the difficulty of discerning \(R_0\) in
\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
model. This model captures the transitions from one state to the next as
a system of ordinary differential equations, where \(N\) is the total
number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

\begin{align}\label{eq:sir-ode}
    S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
    I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
    R^\prime(t) &= \gamma I(t) \nonumber.
\end{align}

From this
model,\textcolor{orange}{the reproduction number is the ratio of the infection rate to the recovery rate,}
\(R_0 = \beta/\gamma\)\sout{, aka the ratio of the infection rate compared to the recovery rate.}
\textcolor{violet}{From this definition, given 
\sout{Since}} \(\beta\) and \(\gamma\) are both rates,
\textcolor{violet}{it should be clear that} the ratio of the two,
\(R_0\), is a time-invariant
quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
\sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.  Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale. }\footnote{cool
  facts about r0, but not the central point}

\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.  also highlighted tie to time-invariant analysis and $R_0$.  I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\textcolor{orange}{ Shannon tries again in \textcolor{blue}{blue}}

\sout{\textcolor{violet}{[Ben says: this paragraph needs to still be looked at. Also I'm not sure why this particular paragraph was changed so much at all. Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations. The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected.  Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.  We plot the epidemics with traditional $state$ vs. $time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
\sout{In Fig. \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.  In this time-variant view, we may believe that \textcolor{violet}{epidemic} 1 has a larger $R_0$ than \textcolor{violet}{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.  On the other hand, we may believe \textcolor{violet}{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \textcolor{violet}{epidemic} has not yet peaked at time 15.  In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.

\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.  Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}

\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$. The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected. An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs. time plots like those shown in Figure \ref{fig:different-scales-standard}. The paths of the epidemics in the state vs. time view seem to differ from one another including having different infection peaks. From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}

@@

\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
\end{figure}
\end{CodeChunk}

\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.}
\sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}
\textcolor{violet}{For every time \sout{point} $t$ we have a \textcolor{orange}{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \textcolor{orange}{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig. \ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right). \sout{In Fig. \ref{fig:different-scales-tern}
we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t
is apparent that these epidemics are on ``the same path.'' In this case,
this indicates that two epidemics have the same value of \(R_0\).

@@

\caption{\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.  Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.  This is more commonly known as a ternary plot.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\textcolor{violet}{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig. \ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space. \sout{The reason why we can visually compare $R_0$ in Fig. \ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}}
A filamental trajectory can be mathematically viewed as a set of points
in space that have an ordering, and that all points on the line between
these ordered points are also contained in the geometric object. For a
SIR epidemic, we can represent the associated filamental trajectory
\(\psi\) as

\[
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
\] where a mapping \(\xi : s \to \mathbb{R}\) that is strictly
monotonically increasing would not change the definition of \(\psi\),
i.e.~\(\psi_\xi \equiv \psi\) where :

\[
\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
\] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
number \textcolor{orange}{of individuals} in each state is non-negative
and the sum over the three states for a given time point
\sout{sums to}\textcolor{orange}{is} \(N\), then
\textcolor{violet}{all points in} \(\psi\) will lay in a
\textbackslash ben\{two-dimensional triangular plane in
three-dimensional space.
\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
As a result, we can visualize
\textcolor{orange}{the full filamental trajectory} of an epidemic in a
single
\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}

\sout{\textcolor{violet}{[This section could be a bit less wordy. But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.  Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.  Now, it is much clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but is not as far along in the infection process. }

\sout{\textcolor{violet}{As suggested a few paragraphs back, t}}The
filamental trajectories in Fig. \ref{fig:different-scales-tern}
\sout{seem to overlap}, and we may suspect that something is
fundamentally linking these two different epidemics together.
Mathematically, we can show this fundamental link \sout{turns} is
\(R_0\). Let our two epidemics be presented as
\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
example, assume both models have the same initial values
\((S(0), I(0), R(0))\), and let
\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
\(\gamma_2 = a \gamma_1\).

\begin{theorem}\label{thm:sir-scale}
Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
point on the epidemic path of the second SIR model.
\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory. For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}

\subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
and \sout{Kermack’s and McKendrick} SIR
Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space allows us
to \sout{better}\textcolor{orange}{more fully} compare the overall
structure of the epidemic and see how the population is directly
impacted. Time-invariant tools \sout{that} can be useful even when the
underlying generative model for the epidemic is unknown or have more
than three epidemic states.

\textcolor{violet}{New paragraph} Viewing epidemics as filamental
trajectories provides \sout{a lot }new ways to compare and examine
epidemics in a time-invariant manner. \textbackslash ben\{For
\textcolor{orange}{completed?}epidemics that have ended, one way to
examine their filamental trajectories is to
\sout{redefine}\textcolor{orange}{represent} the
\textcolor{orange}{filamental} trajectory as a
\textcolor{orange}{finite sequence of equally spaced points.}
\sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).}
\sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
This representation induces a natural distance between this type of
representation between epidemics, specifically: \[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
fraction of \(|\psi_1|\) distance away from the start of
\(\psi_i\).\footnote{\textcolor{orange}{I think you're trying to say something about a distance based on the equally space points.  Some clarifying questions:  is $\psi^\prime$ the derivative?  Does proportion make more sense than fraction? or simply $\frac{|\psi|}{s}$? It's only naturally time-invariant if we have a well defined ending point, right?}}
This distance is naturally time-invariant, and can be plugged into
multiple distance-based assessment tools to examine the overall
``extremeness'\,' of points, including pseudo-density estimators and
depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be \sout{very} useful when comparing
\sout{between a setof simulation} \textcolor{orange}{a set of simulated}
epidemics and the true
epidemic\textcolor{orange}{. Moreover these extremness estimators,}
\sout{and does} \textcolor{orange}{} not constrain the number of states
of the
model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before... \textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
